---
title: "218X Final Assignment"
author: "Merrick Howarth"
date: "Dec. 9, 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
```

# Greenspace and Mental Health


Many studies have shown the benefits of nature exposure on mental health through a variety of indicators which primarily fall into two categories: direct exposure to nature and indirect exposure to nature. Direct exposure to nature involves time people choose to spend outside. Indirect exposure involves time spent near nature, be it in the form of house plants or larger scale impactors like nearby park area. In my research this quarter, I have spent time using park features of tracts as an indicator for indirect exposure on mental health. While my research considers tracts nationally, below I conduct a case study analysis of park prevelance in tracts in San Francisco to evaluate the effectiveness of the indicator and to see if equity impacts this indicator. 


## Mapping the Data

The data which I am considering features park information as recent as 2018, but does in turn seem to rely on 2010 tracts. 

While I have selected 2019 American Community Survey data for my equity analysis, I wanted to confirm that if data was applied on more recent tracts, there was no potential for boundary issues. As shown below, I mapped both 2010 and 2020 census tracts, in blue and red respectively, and on visual inspection, the boundaries seem the same. This is affirmed by numeric evaluation as a comparison of tract areas leads to no differences. 

I additionally felt it was pertinent to give a visual sense of the data regarding park features that I am working with. To this end, I chose to focus on the proportion of tract area which is park both in my analysis and visualization which can be seen in the other map layer below. 


```{r, libraries}

library(tidyverse)
library(plotly)
library(sf)
library(tigris)
library(leaflet)
library(censusapi)

library(mapview)

Sys.setenv(CENSUS_KEY="dbcdeeed2f9e907f31811ee78c1f239a2aa77934")


```

```{r, geospatialdata, include = F}

sf_tracts_10 <- tracts("CA", "San Francisco", 2010) # using 2010 tracts b/c 2018 park data is in 2010 tracts

# park data from https://www.openicpsr.org/openicpsr/project/117921/version/V1/view;jsessionid=C76AD809A54A43E205B0ACC31D208C39
park_data_10 <- read_csv("data/nanda_parks_tract_2018_01P.csv") %>% 
  rename(propPark = prop_park_area_tract,
         GEOID = tract_fips10) 

# joinin parks and tract data by tract id
sf_parks <- left_join(sf_tracts_10,park_data_10) %>% rename(tract = TRACTCE)


# identifying park Proportion quartiles for san fracisco tracts
propPark_Qs <- quantile(sf_parks$propPark, na.rm = T)

sf_parks <- sf_parks %>%
  mutate(
    propPark_Q = case_when(
      propPark > propPark_Qs[4] ~ 4,
      propPark > propPark_Qs[3] ~ 3,
      propPark > propPark_Qs[2] ~ 2,
      propPark >= propPark_Qs[1] ~ 1
    )
  )



```


```{r, show tracts are same}
sf_tracts_20 <- tracts("CA", "San Francisco", 2020)

absolute_pal <- colorNumeric(
  palette = "RdYlGn",
  domain = c(0,1)
)

leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addPolygons(
    data = sf_tracts_10,
    stroke = T,
    opacity = 0.5,
    fill = F,
    group = "Boundry Comparison"
  ) %>% 
  addPolygons(
    data = sf_tracts_20,
    color = "red",
    stroke = T,
    opacity = 0.5,
    weight = 1,
    fill = F,
    group = "Boundry Comparison"
  )%>% 
  # adding Proportion of park shading
  addPolygons(
    data = sf_parks,
    fillColor = ~absolute_pal(sf_parks$propPark),
    label = ~round(sf_parks$propPark,2),
    color = "grey",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    group = "Proportion of Park"
  )%>%
  addLegend(
    pal = absolute_pal,
    value = c(0,1),
    title = "Proportion of Park \n Area to Tract Area"
  ) %>% 
  addLayersControl(
    baseGroups = c("Boundry Comparison","Proportion of Park"),
    options = layersControlOptions(
      collapsed = F
    ))



area_check <- sf_tracts_10 %>% 
  st_transform(26910) %>% 
  mutate(area_10 = st_area(.)) %>%
  
  left_join(sf_tracts_20 %>%
              st_transform(26910) %>%
              mutate(area_20 = st_area(.)) %>% 
              st_drop_geometry() %>% 
              select(GEOID, area_20)) %>%
  
  mutate(delta = area_20 - area_10)

tract_area_dif <- sum(area_check$delta)
  

```

The map above highlights what may prove to be a significant detriment to using park features contained in tracts as an indicator: failure to consider proximity to other park areas. Tracts which boarder the large Golden Gate Park bear similar to worse park proportions when compared to some park sparse tracts in the Mission District. While this issue primarily only presents itself in consideration of large park areas, like Golden Gate Park, it is not a negligible issue.



## Equity Analysis

After visualizing the park data, I wanted to perform an equity for access to parks with consideration of race and income. For the analysis, I broke the tract park proportions across all of the San Francisco tracts into quartiles in order to analyse equity at essentially four different levels of nature exposure. The analysis plots can be seen below.

```{r, tract_income}

# acs_vars_2019_5yr <-
#   listCensusMetadata(
#     name = "2019/acs/acs5",
#     type = "variables"
#   )
# saveRDS(acs_vars_2019_5yr,"data/acs_vars_2019_5yr.rds")
acs_vars_2019_5yr <- read_rds("data/acs_vars_2019_5yr.rds")


census_race_categories <- 
  c(
    "White Alone",
    "Black or African American",
    "American Indian and \n Alaska Native Alone",
    "Asian Alone",
    "Native Hawaiian and \n Other Pacific Islander Alone",
    "Some Other Race Alone",
    "Two or More Races"
  )

sf_income_race_parkQ <-
  1:7 %>% 
  map_dfr(function(x){
    getCensus(
      name = "acs/acs5",
      vintage = 2019,
    region = "tract:*",
    regionin = "state:06+county:075",
      vars = paste0("group(B19001",LETTERS[x],")")
    ) %>%
      select(!c(GEO_ID,state,NAME,county) & !ends_with(c("EA","MA","M"))) %>%
      pivot_longer(
        ends_with("E"),
        names_to = "name",
        values_to = "Proportion"
      ) %>%
      left_join(
        acs_vars_2019_5yr %>% 
          select(name, label)
      ) %>% 
      select(-name) %>% 
      separate(
        label,
        into = c(NA,NA,"income"),
        sep = "!!"
      ) %>% 
      filter(!is.na(income)) %>% 
      mutate(race = census_race_categories[x])
  }
  ) %>%
  
  left_join(sf_parks %>% select(tract, propPark_Q)) %>%
  
  mutate(propPark_Q = as.character(propPark_Q))


# reordering by quartile for plotting purposes

sf_income_race_parkQ <- sf_income_race_parkQ %>% arrange(desc(propPark_Q))

```


```{r, equity plot race}

sf_race_total <-
  sf_income_race_parkQ %>% 
  group_by(race) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  mutate(income = "Total Pop.", propPark_Q = "Total Pop.") 

sf_park_race_gg <- sf_income_race_parkQ %>% 
  group_by(propPark_Q, race) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  rbind(sf_race_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = propPark_Q %>% factor(levels = rev(c("Total Pop.",unique(sf_income_race_parkQ$propPark_Q)))),
      y = Proportion,
      fill = race %>% factor(levels = rev(unique(sf_income_race_parkQ$race)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Park Quartile",
    y = "Proportion of households",
    title = "San Francisco Park Proportion of Tract by Race",
    fill = "Race of householder"
  ) #+
  #coord_flip() +
  # theme(
  #   legend.position = "bottom",
  #   legend.direction = "vertical"
  # )

sf_park_race <- ggplotly(sf_park_race_gg, tooltip="y") %>% layout(legend = list(x = 200, y = 0.5))

sf_park_race

```


```{r, equity plot income}

sf_income_total <-
  sf_income_race_parkQ %>% 
  group_by(income) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  mutate(race = "Total Pop.", propPark_Q = "Total Pop.") 


sf_park_income_gg <- sf_income_race_parkQ %>% 
  group_by(propPark_Q, income) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  rbind(sf_income_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = propPark_Q %>% factor(levels = rev(c("Total Pop.",unique(sf_income_race_parkQ$propPark_Q)))),
      y = Proportion,
      fill = income %>% factor(levels = rev(unique(sf_income_race_parkQ$income)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Park Quartile",
    y = "Proportion of Households",
    title = "San Francisco Park Proportion of Tract by Household Income",
    fill = "Household Income Bracket"
  ) 

sf_park_income <- ggplotly(sf_park_income_gg, tooltip="y")

sf_park_income

```


```{r, equity plot income}
# mutate to larger buckets for cleaner plot
sf_income_race_parkQ <- sf_income_race_parkQ %>%
  mutate(
    income1 = income,
    income = case_when(
      
      income == "Less than $10,000" | income == "$10,000 to $14,999" |  
        income == "$15,000 to $19,999" 
      ~ "Less than $19,999",
      
      income == "$20,000 to $24,999" | income == "$25,000 to $29,999" |  
        income == "$30,000 to $34,999" | income == "$35,000 to $39,999"
      ~ "$20,000 to $39,999",
      
      income == "$40,000 to $44,999" | income == "$45,000 to $49,999" |  
        income == "$50,000 to $59,999"
      ~ "$40,000 to $59,999",
      
      TRUE ~ income,
      
      
    )
      
  )


sf_income_total <-
  sf_income_race_parkQ %>% 
  group_by(income) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  mutate(race = "Total Pop.", propPark_Q = "Total Pop.") 


sf_park_income_gg <- sf_income_race_parkQ %>% 
  group_by(propPark_Q, income) %>% 
  summarize(Proportion = sum(Proportion)) %>% 
  rbind(sf_income_total) %>% 
  ggplot() +
  geom_bar(
    aes(
      x = propPark_Q %>% factor(levels = rev(c("Total Pop.",unique(sf_income_race_parkQ$propPark_Q)))),
      y = Proportion,
      fill = income %>% factor(levels = rev(unique(sf_income_race_parkQ$income)))
    ),
    stat = "identity",
    position = "fill"
  ) +
  labs(
    x = "Park Quartile",
    y = "Proportion of Households",
    title = "San Francisco Park Proportion of Tract by Household Income",
    fill = "Household Income Bracket"
  ) 

sf_park_income <- ggplotly(sf_park_income_gg, tooltip="y")

sf_park_income

```

Ultimately, it can be easily seen that tract contained park features are an inadequate indicator for capturing park access and an in depth proximity analysis is needed for having a chance at drawing significant conclusions regarding equity and in my own research mental health. This presents a significant challenge as I continue my research as more precise geospatial park data is limited on a national scale. In future analysis, I may create another midpoint indicator which applies a buffer analysis to the tracts themselves, in order to capture the park information of adjacent tract in a larger park access score.




